Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-04-15 pldi-2016/paper.tex
Published version

For example, \url{http://openweathermap.org/current} contains one example to document an end-point 
to get the current weather. Using standard libraries, we might call it as\footnote{We use an abbreviation
\url{http://api.owm.org/?q=X} for the full URL:
\url{http://api.openweathermap.org/data/2.5/weather?q=X} 
}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field representing the current temperature. When the shape is different, 
the code fails. While not immediately unsound, the code is prone to errors if strings are 
misspelled or incorrect shape assumed.

@@

We start with an informal overview that shows how F\# Data type providers simplify working with 
JSON and XML. We introduce the necessary aspects of F\# type providers along the way. The examples 
in this section illustrate the key design principles of the shape inference algorithm:

@@

\item The above also makes our techniques applicable to any language with nominal 
  object types (e.g.~variations of Java or C\# with a type provider mechanism added).

@@

\noindent
The supplementary screencast provides further illustration of the practical developer
experience using F\# Data.

@@

Assuming \strf{people.json} is the above example and \ident{data} is a string containing
JSON of the same shape, we can write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
In contrast to the earlier example, we now use a local file \strf{people.json} as a sample for 
the type inference, but then processes data from another source. The code achieves a similar 
simplicity as when using dynamically typed languages, but it is statically type-checked.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and provides a type  \ident{People}. F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion on ``.'' and background type-checking.

The \ident{JsonProvider} uses a shape inference algorithm and provides
the following F\# types for the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.5em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. In F\#, we can then use \ident{Option.iter} to call
the specified function (printing) only when an optional value is available. The two age values 
are an integer $25$ and a float $3.5$ and so the common inferred type is \ident{float}.
The names of the properties are normalized to follow standard F\# naming conventions
as discussed later (\S\ref{sec:impl-naming}).

@@

\paragraph{The role of objects with properties.}
The sample code is easy to write thanks to the fact that most F\# editors provide auto-completion
when ``.'' is typed (see the supplementary screencast). The developer does not need to examine the 
sample JSON file to see what fields are available. To support this scenario, our type providers 
map the inferred shapes to F\# objects with (possibly optional) properties.

This is demonstrated by the fact that \ident{Age} becomes an optional member.
An alternative is to provide two different record types (one with \ident{Name} and other with 
\ident{Name} and \ident{Age}), but this would complicate the processing code.
It is worth noting that languages with stronger tooling around pattern matching
such as Idris \cite{idris-tools} might have different preferences.

@@

Consider a simple extensible document format where a root element {\ttfamily\small <doc>} can 
contain a number of document elements, one of which is {\ttfamily\small <heading>} representing 
headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarily to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
provides typed access to elements known statically from the sample and so we can write \ident{elem.Heading}, 
which returns an optional string value.

@@

Throughout the introduction, we used data sets that demonstrate the typical problems frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
The government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three:
%
{\small{
\begin{verbatim}
  [ { "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First, the field \ident{value} is \kvd{null} for some records. Second, numbers in JSON can be 
represented as numeric literals (without quotes), but here, they are returned as string literals 
instead.\footnote{This is often used to avoid non-standard numerical types of JavaScript.}
Finally, the top-level element is a collection containing two values of different shape.
The record contains meta-data with the total number of pages and the array contains the data.
F\# Data supports a concept of heterogeneous collection (outlined in in \S\ref{sec:impl-hetero})
and provides the following type:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{Record}~=  \\
 \quad \kvd{member}~\ident{Pages}~:~\ident{int} \\[0.5em]
\kvd{type}~\ident{Item}~=  \\
 \quad \kvd{member}~\ident{Date}~:~\ident{int} \\
 \quad \kvd{member}~\ident{Indicator}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Value}~:~\ident{option}\langl \ident{float}\rangl\\[0.5em]
\kvd{type}~\ident{WorldBank}~=  \\
 \quad \kvd{member}~\ident{Record}~:~\ident{Record}\\
 \quad \kvd{member}~\ident{Array}~:~\ident{Item}[]\\[0.5em]
\end{array}
\end{equation*}
%
The inference for heterogeneous collections infers the multiplicities and shapes of nested 
elements. As there is exactly one record and one array, the provided type \ident{WorldBank} exposes 
them as properties \ident{Record} and \ident{Array}.

@@

The shape algebra and inference presented here is influenced by the design principles
we outlined earlier and the type definitions available in the F\# language.
If applied to other languages, details may differ, for example
with respect to numerical types and missing data.

@@

During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent 
the flexibility arising from records in samples. For example, when a 
record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
overall shape inferred must account for the fact that any extra fields are optional,
giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{nullable}\langl\ident{int}\rangl\}$.

@@

\noindent
\begin{align}
  \label{eq:sub-prim}\tag{P1}
  \ident{int}\,&\sqsubseteq\,\ident{float}&\\[-0.2em]
  \label{eq:sub-null}\tag{P2}
  \kvd{null} &\sqsubseteq \sigma  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}\tag{P3}
  \hat{\sigma} &\sqsubseteq \kvd{nullable}\langl\hat{\sigma}\rangl  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}\tag{P4}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsubseteq 
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsubseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}\tag{P5}
  [\sigma_1] &\sqsubseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsubseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}\tag{P6}
  \bot &\sqsubseteq \sigma  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}\tag{P7}
  \sigma &\sqsubseteq \kvd{any}
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}\tag{R1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsubseteq \sigma_i')
\end{align}

@@

\begin{align}
\label{eq:sub-record2}\tag{R2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \leq n)
\end{align}

@@

Primitive  values are mapped to their corresponding shapes. 
When inferring a shape from multiple samples, we use the common preferred shape relation to find a 
common shape for all values (starting with $\bot$). This operation is used  
when calling a type provider with multiple samples and also when inferring the shape of collection
elements.

@@

When combining two top shapes (\emph{top-merge}), we group the annotations by their tags.
When combining a top with another shape, the labels may or may not already contain a case with 
the tag of the other shape. If they doe, the two shapes are combined (\emph{top-incl}), otherwise 
a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any}) 
and combines two distinct non-top shapes. As top shapes implicitly permit \kvd{null} values,
we make the labels non-nullable using $\dropopt{-}$.

@@

\paragraph{Reduction.}
The reduction relation is of the form $L, e \reduce e'$. We omit class declarations 
$L$ where implied by context and write $e \reduce^{*} e'$ for
the reflexive, transitive closure of $\reduce$.

@@

All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
implementation provides \emph{lazy erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
do not rely on laziness or erasure of type provision.}

@@

For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
record field, we generate a member with the same name as the field. The body of the member calls 
\ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field 
value (data of shape $\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
the class name $C$ is not directly accessed by the user and so we can use an arbitrary name, although the 
actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
based on the name of the parent record field.}

@@

\paragraph{Example 1.}
To illustrate how the mechanism works, we consider two examples. First, assume 
that the inferred shape is a record  
$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
The rules from Figure~\ref{fig:tp-generation} produce the following class \ident{Person} (shown below)
with two members.

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
\item[--] Records in the input can have additional fields
\item[--] Records in the input can have fewer fields than some of the records in the sample 
  document, provided that the sample also contains records that do not have the field 
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption  
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

\begin{lemma}[Correctness of provided types]
\label{thm:tp-correctness}
Given sample data $d$ and an input data value $d'$ such that $\semalt{d'} \sqsubseteq \semalt{d}$
and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$, 
then $L, e~d' \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
class $C$, it holds that $L, (e~d').N_i \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subtypes.
For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
works on \kvd{null} values, \ident{convFloat} accepts an integer). Finally, for labelled top types,
the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
\end{proof}

@@

The F\# Data library has been widely adopted by users and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 100,000 downloads on NuGet 
(package repository), 1,844 commits and 44 contributors on GitHub.} A practical demonstration of 
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relative safety} property does not guarantee safety in the same way as traditional
closed-world type safety, but it reflects the reality that programming with external data is
increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of 
programming.

@@

\paragraph{Richer data sources.}
In general, XML, CSV and JSON data sources without explicit schema will necessarily require 
techniques akin to those we have shown. However, some data sources provide explicit schema with 
schema versioning support. For those, a type provider could be written to adapt automatically, 
but we leave this for future work.

@@

In our formalization, we treat XML, JSON and CSV uniformly as \emph{data values}. With the addition of 
names for records (for XML nodes), the definition of structural values is rich enough to capture all 
three formats\footnote{The same mechanism has later been used by the HTML type provider 
(\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy 
access to data in HTML tables and lists.}. However, parsing real-world data poses a number of practical issues.

@@

Our current implementation also adds an additional member to each class that returns the
underlying JSON node (called \ident{JsonValue}) or XML element (called \ident{XElement}). 
Those return standard .NET or F\# representation of the value and can be used to dynamically 
access data not exposed by the type providers, such as textual values inside mixed-content XML elements.

@@

The work that is most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
It extends C\# language with types similar to our structural types 
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).

>>>>

For example, \url{http://openweathermap.org/current} contains one example to document an end-point
to get the current weather. Using standard libraries, we might call it as\footnote{We abbreviate
the full URL and omit application key (available after registration). The returned JSON is shown
in Appendix~\ref{sec:appendix-weather} and can be used to run the code against a local file.}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format}
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field representing the current temperature. When the shape is different,
the code fails. While not immediately unsound, the code is prone to errors if strings are
misspelled or incorrect shape assumed.

@@

We start with an informal overview that shows how F\# Data type providers simplify working with
JSON and XML. We introduce the necessary aspects of F\# type providers along the way. The examples
in this section also illustrate the key design principles of the shape inference algorithm:

@@

\item The above makes our techniques applicable to any language with nominal
  object types (e.g.~variations of Java or C\# with a type provider mechanism added).

@@

\noindent
The supplementary screencast provides further illustration of the practical developer
experience using F\# Data.\footnote{Available at \url{http://tomasp.net/academic/papers/fsharp-data}.}

@@

Assuming \strf{people.json} is the above example and \ident{data} is a string containing
JSON of the same shape, we can write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
 \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
 \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
\end{array}
\end{equation*}
%
We now use a local file as a sample for
the type inference, but then processes data from another source. The code achieves a similar
simplicity as when using dynamically typed languages, but it is statically type-checked.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter}
to the type provider. Static parameters are resolved at compile-time and have to be constant.
The provider analyzes the sample and provides a type  \ident{People}. F\# editors also
execute the type provider at development-time and use the provided types
for auto-completion on ``.'' and for background type-checking.

The \ident{JsonProvider} uses a shape inference algorithm and provides
the following F\# types for the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\[-0.05em]
 \quad \kvd{member}~\ident{Name}~:~\ident{string} \\[-0.05em]
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.4em]
 \kvd{type}~\ident{People}~=  \\[-0.05em]
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\[-0.05em]
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\[-0.05em]
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. In F\#, we use \ident{Option.iter} to call
the specified function (printing) only when an optional value is available. The two age values
are an integer $25$ and a float $3.5$ and so the common inferred type is \ident{float}.
The names of the properties are normalized to follow standard F\# naming conventions
as discussed later (\S\ref{sec:impl-naming}).

@@

\paragraph{Objects with properties.}
The sample code is easy to write thanks to the fact that most F\# editors provide auto-completion
when ``.'' is typed (see the supplementary screencast). The developer does not need to examine the
sample JSON file to see what fields are available. To support this scenario, our type providers
map the inferred shapes to F\# objects with (possibly optional) properties.

This is demonstrated by the fact that \ident{Age} becomes an optional member.
An alternative is to provide two different record types (one with \ident{Name} and one with
\ident{Name} and \ident{Age}), but this would complicate the processing code.
It is worth noting that languages with stronger tooling around pattern matching
such as Idris \cite{idris-tools} might have different preferences.

@@

Consider a simple extensible document format where a root element {\ttfamily\small <doc>} can
contain a number of document elements, one of which is {\ttfamily\small <heading>} representing
headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarily to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{pldi/another.xml})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem}
provides typed access to elements known statically from the sample and so we can write \ident{elem.Heading},
which returns an optional string value.

@@

Throughout the introduction, we used data sets that demonstrate the typical problems frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes).
The government debt information returned by the World Bank\footnote{Available at
\url{http://data.worldbank.org}} includes all three:
%
{\small{
\begin{verbatim}
  [ { "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First, the field \ident{value} is \kvd{null} for some records. Second, numbers in JSON can be
represented as numeric literals (without quotes), but here, they are returned as string literals
instead.\footnote{This is often used to avoid non-standard numerical types of JavaScript.}
Finally, the top-level element is a collection containing two values of different shape.
The record contains meta-data with the total number of pages and the array contains the data.
F\# Data supports a concept of heterogeneous collection (outlined in in \S\ref{sec:impl-hetero})
and provides the following type:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{Record}~=  \\
 \quad \kvd{member}~\ident{Pages}~:~\ident{int} \\[0.5em]
\kvd{type}~\ident{Item}~=  \\
 \quad \kvd{member}~\ident{Date}~:~\ident{int} \\
 \quad \kvd{member}~\ident{Indicator}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Value}~:~\ident{option}\langl \ident{float}\rangl\\[0.5em]
\kvd{type}~\ident{WorldBank}~=  \\
 \quad \kvd{member}~\ident{Record}~:~\ident{Record}\\
 \quad \kvd{member}~\ident{Array}~:~\ident{Item}[]\\[0.5em]
\end{array}
\end{equation*}
%
The inference for heterogeneous collections infers the multiplicities and shapes of nested
elements. As there is exactly one record and one array, the provided type \ident{WorldBank} exposes
them as properties \ident{Record} and \ident{Array}.

@@

The shape algebra and inference presented here is influenced by the design principles
we outlined earlier and by the type definitions available in the F\# language.
The same principles apply to other languages, but details may differ, for example
with respect to numerical types and missing data.

@@

During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent
the flexibility arising from records in samples. For example, when a
record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contains more fields. The
overall shape inferred must account for the fact that any extra fields are optional,
giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{nullable}\langl\ident{int}\rangl\}$.

@@

\noindent
\begin{align}
  \label{eq:sub-prim}
  \ident{int}\,&\sqsubseteq\,\ident{float}&\\[-0.2em]
  \label{eq:sub-null}
  \kvd{null} &\sqsubseteq \sigma  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
  \label{eq:sub-opt}
  \hat{\sigma} &\sqsubseteq \kvd{nullable}\langl\hat{\sigma}\rangl  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
  \label{eq:sub-opt-cov}
  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsubseteq
    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsubseteq \hat{\sigma_2})\\[-0.2em]
  \label{eq:sub-col}
  [\sigma_1] &\sqsubseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsubseteq \sigma_2) \\[-0.2em]
  \label{eq:sub-bot}
  \bot &\sqsubseteq \sigma  &(\textnormal{for all}~\sigma)\\[-0.2em]
  \label{eq:sub-var-top}
  \sigma &\sqsubseteq \kvd{any}
\end{align}
\vspace{-2em}

\noindent
\begin{align}
\label{eq:sub-record1}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsubseteq \sigma_i')
\end{align}

@@

\begin{align}
\label{eq:sub-record2}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
 \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
\end{array} \quad~~\, (\textnormal{when}~m \leq n)
\end{align}

@@

Primitive  values are mapped to their corresponding shapes.
When inferring a shape from multiple samples, we use the common preferred shape relation to find a
common shape for all values (starting with $\bot$). This operation is used
when calling a type provider with multiple samples and also when inferring the shape of collection
values.

@@

When combining two top shapes (\emph{top-merge}), we group the annotations by their tags.
When combining a top with another shape, the labels may or may not already contain a case with
the tag of the other shape. If they do, the two shapes are combined (\emph{top-incl}), otherwise
a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any})
and combines two distinct non-top shapes. As top shapes implicitly permit \kvd{null} values,
we make the labels non-nullable using $\dropopt{-}$.

@@

\paragraph{Reduction.}
The reduction relation is of the form $L, e \reduce e'$. We omit class declarations
$L$ where implied by the context and write $e \reduce^{*} e'$ for
the reflexive, transitive closure of $\reduce$.

@@

All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual
implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat
the code as actually generated. This is an acceptable simplification, because F\# Data type providers
do not rely on laziness or erasure of type provision.}

@@

For records, we generate a class $C$ that takes a data value as a constructor parameter. For each
field, we generate a member with the same name as the field. The body of the member calls
\ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field
value (data of shape $\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of
$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that
the class name $C$ is not directly accessed by the user and so we can use an arbitrary name, although the
actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in
\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
based on the name of the parent record field.}

@@

\paragraph{Example 1.}
To illustrate how the mechanism works, we consider two examples. First, assume
that the inferred shape is a record
$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl,~\ident{Name}\!:\!\ident{string}~ \}$.
The rules from Figure~\ref{fig:tp-generation} produce the \ident{Person} class shown below
with two members.

@@

More formally, given a set of sample documents, code using the provided type is guaranteed to work if
the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to
\S\ref{sec:inference-subtyping}, this means that:
%
\begin{itemize}
\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer).
\item[--] Records in the input can have additional fields.
\item[--] Records in the input can have fewer fields than some of the records in the sample
  document, provided that the sample also contains records that do not have the field.
\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
  which implements the open world assumption.
\end{itemize}
%
The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
expression (with input $d'$) can be reduced to a value and, if it is a class,
all its members can also be reduced to values.

\begin{lemma}[Correctness of provided types]
\label{thm:tp-correctness}
Given sample data $d$ and an input data value $d'$ such that $\semalt{d'} \sqsubseteq \semalt{d}$
and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$,
then $L, e~d' \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the
class $C$, it holds that $L, (e~d').N_i \reduce^{*} v$.
\end{lemma}
\begin{proof}
By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subshapes.
For other cases, analyze the provided code to see that it can work on all subshapes (for example~\ident{convElements}
works on \kvd{null} values, \ident{convFloat} accepts an integer). Finally, for labelled top types,
the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
\end{proof}

@@

The F\# Data library has been widely adopted by users and is one of the most downloaded
F\# libraries.\footnote{At the time of writing, the library has over 125,000 downloads on NuGet
(package repository), 1,844 commits and 44 contributors on GitHub.} A practical demonstration of
development using the library can be seen in an attached screencast and additional documentation can be
found at \url{http://fsharp.github.io/FSharp.Data}.

@@

The \emph{relative safety} property does not guarantee safety in the same way as traditional
closed-world type safety, but it reflects the reality of programming with external data that is
becoming increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of
programming.

@@

\paragraph{Richer data sources.}
In general, XML, CSV and JSON data sources without an explicit schema will necessarily require
techniques akin to those we have shown. However, some data sources provide an explicit schema with
versioning support. For those, a type provider that adapts automatically could be written,
but we leave this for future work.

@@

In our formalization, we treat XML, JSON and CSV uniformly as \emph{data values}. With the addition of
names for records (for XML nodes), the definition of structural values is rich enough to capture all
three formats.\footnote{The same mechanism has later been used by the HTML type provider
(\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy
access to data in HTML tables and lists.} However, parsing real-world data poses a number of practical issues.

@@

Our current implementation also adds an additional member to each class that returns the
underlying JSON node (called \ident{JsonValue}) or XML element (called \ident{XElement}).
Those return the standard .NET or F\# representation of the value and can be used to dynamically
access data not exposed by the type providers, such as textual values inside mixed-content XML elements.

@@

The work that is the most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
It extends C\# language with types similar to our structural types
(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
However, C$\omega$ does not infer the types from samples and extends the type system of the host
language (rather than using a general purpose embedding mechanism).
